Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\

### 2025-05-27 paper/main.tex
advancements

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{physics}

@@

\title{Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\
{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured for https://ieeexplore.ieee.org  and
should not be used}
}

@@

\begin{equation}
    \begin{aligned}
        H(t) &= \dfrac{\Omega(t)}{2}\sum_j \left(\ket{g_j}\bra{r_j}+\ket{r_j}\bra{g_j}\right) \\
             &{{\quad}} + \sum_{j<k}V_{jk}n_jn_k - \sum_j \left[\Delta_{\mathrm{g}}(t) + \alpha_j\Delta_{\mathrm{l}}(t)\right] n_j,
    \end{aligned}
\end{equation}

@@

This typical architecture of an autoencoder is illustrated in 
`Fig.-\ref{fig:autoencoder_architecture}'.
\input{autoencoder_architecture.tex}

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}
Our proposed pipeline integrates classical and quantum 
components to perform dimensionality reduction and
classification. 
A classical autoencoder compresses
high-dimensional image data into a lower-dimensional
latent space. 
This compressed representation is 
encoded into a Rydberg Atom chain, which serves as 
the input to a quantum reservoir. 
The reservoir will then produce quantum embeddings
by measuring quantum observables, which capture the
dynamics of the quantum system in response to the input data.

These quantum embeddings will be used to train a
surrogate model that aims to 
approximates the quantum reservoir's behaviour.

The surrogate model's outputs will then pass 
through a linear classifier, which is trained
to map the approximated quantume embeddings to class labels.

The autoencoder is trained to minimize in interpolation
between the data reconstruction's loss function as well 
as the classification task loss function.

Our approach, allows the model to learn representations that are not only
compact and informative for the reconstruction of the input data,
but also well-suited for quantum reservoir computing.

When autoencoder model is trained,
we can use the trained encoder to compress the train
data into the latent space, and then pass the 
compressed data through the real quantum reservoir.
The obtained embeddings are then used to train a linear
classifier that maps the quantum embeddings to class labels.

@@

% input pdf image
\begin{figure*}[!b]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \includegraphics{qgars_pipeline_cropped.pdf}
    }
    \label{fig:qgars_pipeline}
    \caption{Overview of the Quantum-Guided Autoencoder Reservoir Computing System.}
\end{figure*}

@@

\subsubsection{Loss Function Design}
% Loss Function Design

% The QGA is trained to minimize a composite loss function that balances reconstruction fidelity and classification accuracy:
% L=λrec⋅Lrec+λcls⋅Lcls,
% L=λrec​⋅Lrec​+λcls​⋅Lcls​,

% where:

%     Lrec=∥x−x^∥2Lrec​=∥x−x^∥2 is the reconstruction loss, measuring the mean squared error between the input xx and its reconstruction x^x^.

%     Lcls=−∑iyilog⁡(y^i)Lcls​=−∑i​yi​log(y^​i​) is the classification loss, computed as the cross-entropy between the true labels yiyi​ and the predicted probabilities y^iy^​i​.

%     λrecλrec​ and λclsλcls​ are hyperparameters that control the trade-off between reconstruction and classification objectives [2].

The QGA is trained to minimize a composite loss 
function that balances reconstruction fidelity and 
classification accuracy:

@@

\subsubsection{Balancing Reconstruction and Classification}

\subsection{The Gradient Barrier Problem}
\subsection{Surrogate Modeling for Quantum Layers}
\subsubsection{Architecture and Training}
\subsubsection{Gradient Flow Through Surrogate Models}
\subsection{Rydberg Hamiltonian and Quantum Dynamics}
\subsection{Data Encoding Schemes}
\subsection{Quantum Readout Methods}
\subsubsection{Single-atom Measurements}
\subsubsection{Two-atom Correlations}
\subsubsection{Three-atom Correlations}

@@

%============================================
% THEORETICAL ANALYSIS
%============================================
\section{Theoretical Analysis}
\subsection{Information Encoding in Quantum Reservoirs}
\subsection{Gradient Flow in Quantum-Classical Hybrid Systems}
\subsection{Computational Complexity}
\subsection{Quantum Resource Requirements}

>>>>

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{float}

@@

\title{Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\
% {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured for https://ieeexplore.ieee.org  and
% should not be used}
}

@@

\begin{equation}
    \begin{aligned}
        H(t) &= \dfrac{\Omega(t)}{2}\sum_j \left(\ket{g_j}\bra{r_j}+\ket{r_j}\bra{g_j}\right) \\
             &{{\quad}} + \sum_{j<k}V_{jk}n_jn_k - \sum_j \left[\Delta_{\mathrm{g}}(t) + \alpha_j\Delta_{\mathrm{l}}(t)\right] n_j,
    \end{aligned}
    \label{eq:rydberg_hamiltonian}
\end{equation}

@@

This typical architecture of an autoencoder is illustrated in 
`Fig.-\ref{fig:autoencoder_architecture}'.

\input{autoencoder_architecture.tex}

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}
Our proposed pipeline integrates classical and quantum 
components for dimensionality reduction and classification. 
A classical autoencoder compresses high-dimensional image 
data into a lower-dimensional latent space. This latent 
representation is encoded into a Rydberg atom chain, 
serving as input to a quantum reservoir. The reservoir's 
dynamics, governed by the Rydberg Hamiltonian, produce 
quantum embeddings through measurements of quantum observables.

To facilitate end-to-end training, we employ a differentiable 
surrogate model that approximates the quantum reservoir's 
behavior, enabling gradient flow from the classification 
loss back to the autoencoder. The surrogate's outputs feed 
into a linear classifier trained to map quantum embeddings to class labels.

The autoencoder is trained to minimize a combined loss 
function that interpolates between reconstruction loss and 
classification loss, promoting representations that are both 
informative for input reconstruction and effective for quantum 
reservoir computing.

Once trained, the encoder compresses data into the latent 
space, which is then processed by the actual quantum reservoir. 
The resulting quantum embeddings are used to train a linear 
classifier for final predictions.

@@

\begin{figure*}[!h]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \includegraphics{qgars_pipeline_cropped.pdf}
    }
    \label{fig:qgars_pipeline}
    \caption{Overview of the Quantum-Guided Autoencoder Reservoir Computing System.}
\end{figure*}

@@

\subsubsection{Loss Function Design}
The QGA is trained to minimize a composite loss 
function that balances reconstruction fidelity and 
classification accuracy:

@@

%\subsubsection{Balancing Reconstruction and Classification}

\subsection{The Gradient Barrier Problem}
Hybrid quantum-classical models, such as the Quantum Guided Autoencoder 
(QGAE), aim to leverage quantum computational advantages within 
classical machine learning frameworks. However, a significant 
challenge arises due to the non-differentiable nature of quantum 
operations, commonly referred to as the `gradient barrier'.

In classical neural networks, training relies on backpropagation, 
which requires the computation of gradients through all components 
of the network. In hybrid models, the inclusion of quantum 
layers introduces operations that are inherently non-differentiable:

\begin{itemize}
    \item \textbf{Quantum State Preparation}: 
    Encoding classical data into 
    quantum states often involves operations 
    that are not smoothly differentiable with 
    respect to the input data.

    \item \textbf{Quantum Measurements}:
    Observing quantum states collapses them, 
    introducing stochasticity and breaking the 
    deterministic gradient flow required 
    for backpropagation.
\end{itemize}

These aspects hinder the direct application of 
gradient-based optimization methods across the 
quantum-classical boundary, obstructing end-to-end 
training of hybrid models.

\subsection{Surrogate Modeling for Quantum Layers}
To circumvent the gradient barrier, surrogate models 
have been proposed. These are classical, differentiable 
models trained to approximate the input-output behavior 
of quantum circuits. By replacing the non-differentiable 
quantum components with their surrogate counterparts during 
training, gradients can be propagated through the entire network, 
enabling end-to-end optimization.

Recent studies have demonstrated the efficacy of 
surrogate models in mitigating the effects of barren plateaus 
and facilitating the training of hybrid quantum-classical models. 
These approaches allow for the practical implementation of 
quantum-enhanced machine learning models by leveraging classical 
optimization techniques while still capturing quantum computational 
advantages~\cite{xieQuantumSurrogateDrivenImage2025b}

\subsubsection{Architecture}
In our pipeline, the surrogate model is implemented as a 
feedforward neural network designed to learn the mapping 
from the autoencoder's latent space to the quantum 
embeddings produced by the quantum reservoir.

Its architecture comprises multiple fully connected 
layers with nonlinear activation functions, facilitating 
the approximation of the complex, non-linear 
transformations inherent in quantum dynamics.

This approach draws inspiration from recent studies 
demonstrating the effectiveness of surrogate models 
in approximating quantum circuit behaviors, thereby 
facilitating efficient training of hybrid 
models.~\cite{xieQuantumSurrogateDrivenImage2025b,schreiberClassicalSurrogatesQuantum2023}

\subsubsection{Training Procedure}
\begin{itemize}
    \item Data Input: 
    Feed a batch of latent data \( z \) from the autoencoder 
    into the real quantum reservoir, recording the resulting quantum 
    embeddings.

    \item Surrogate Training:
    Train the surrogate model using the recorded quantum
    embeddings as targets, minimizing the loss function:
    \begin{equation}
        \mathcal{L}_{\text{surrogate}} = \| \hat{y} - y \|^2,
    \end{equation}
    where \( \hat{y} \) are the surrogate model's predictions
    and \( y \) are the actual quantum embeddings.

    \item Regular Updates:
    Periodically update the surrogate model during training
    to insure it accurately reflects the quantum reservoir's
    behavior. This can be done after a fixed number of epochs
    or when the surrogate's performance on a validation set
    reaches a certain threshold.
\end{itemize}

\subsubsection{Gradient Flow Through Surrogate Models}
By integrating the surrogate model into the training pipeline, 
we establish a continuous computational graph from the 
output layer back to the input layer, enabling the use of 
gradient-based optimization techniques. The surrogate model 
serves as a differentiable proxy for the quantum reservoir, 
allowing the classification loss to influence the 
autoencoder's parameters effectively.

As a results, we not only achieve end-to-end training of the
hybrid model but also makes it computationally efficient,
as we reduce the need for frequent quantum measurements.

\subsection{Rydberg Hamiltonian and Quantum Dynamics}
The quantum reservoir in our architecture is implemented
using a neutral-atom system, specifically utilizing
Rydberg atoms to create a programmable quantum reservoir.
The dynamics of the system are governed by the Rydberg Hamiltonian,
which describes the interactions between atoms in Rydberg states~\ref{eq:rydberg_hamiltonian}.

\subsubsection{Data Encoding Schemes}
The input data is encoded into the quantum system by mapping
the latent representations from the autoencoder to the
detuning parameters of the Rydberg Hamiltonian.
This encoding process involves adjusting the detuning
parameters \( \Delta_{\mathrm{l}}(t) \) for each atom in the chain,
allowing the system to represent the input data in a way that
is compatible with the quantum dynamics of the reservoir.

\subsubsection{Quantum Readout Methods}
The quantum reservoir's output is obtained by probing, 
in successive time steps, the state of the atoms 
after the systems evolution up to a certain time \( t \).

The readout is performed by measuring specific observables,
such as:
\begin{itemize}
    \item \textbf{Single-atom Measurements}: 
    \( \langle Z_i \rangle \) for each atom \( i \), where \( Z_i \) is the Pauli-Z operator.
    \item \textbf{Two-atom Correlations}: 
    \( \langle Z_i Z_j \rangle \) for pairs of atoms \( (i, j) \), capturing correlations between their states.
    \item \textbf{Three-atom Correlations}:
    \( \langle Z_i Z_j Z_k \rangle \) for triplets of atoms \( (i, j, k) \), providing higher-order correlations.
\end{itemize}

@@
